---
title: "California Spiny Lobster Habitat Assessment for Aquaculture Opportunities"
author: "Karol Paya"
date: Dec. 7, 2024
execute: 
  eval: True
  warning: False
  message: False
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---
## Study Background
This study employs a geospatial analysis to assess the suitability of habitats for spiny lobsters (Panulirus interruptus) in California, with a particular focus on identifying areas most conducive to aquaculture development. Spiny lobsters are an economically significant species in the region, and their population dynamics are influenced by factors such as temperature and ocean depth. However, traditional fishing practices have led to pressure on wild populations, prompting a growing interest in sustainable aquaculture as a potential solution. By remote sensing technologies, this analysis integrates environmental data, such as sea surface temperature and Bathymetry, to search for potential areas that could support viable lobster farming operations. 

## Data Details

**Suitable growing conditions**

Using California spiny lobster *(Panulirus interruptus)* to showcase the suitable habitat function

Location: Eastern Central Pacific, from San Luis Obispo Bay, California, USA, to Magdalena Bay, west coast of Baja California Sur, and along the west coast of the Gulf of California, to Carmen Island, Baja California Sur, Mexico.

Research has shown that lobsters need the following conditions for optimal growth:
- Preferred temperature: 14.8c - 22.3c
- Preferred depth: 0 - 150 m 

Source: https://www.sealifebase.ca/summary/Panulirus-interruptus.html

**Sea Surface Temperature**

National Oceanic and Atmospheric Administration (NOAA). 2008-2012. 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1 [Data set]. NOAA National Centers for Environmental Information. Retrieved Nov. 14, 2024 from https://www.ncei.noaa.gov

**Bathymetry**

General Bathymetric Chart of the Oceans (GEBCO). 2024. gridded bathymetry data [Data set]. Retrieved Nov. 14, 2024 from https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area

**Exclusive Economic Zones**

MarineRegions.org. (n.d.). Exclusive Economic Zones (EEZs) off the west coast of the United States [Data set]. Retrieved Nov. 14, 2024 fromhttps://www.marineregions.org/

## Load Libraries

<details>
<summary>Click to view code</summary>
```{r}
# Load libraries
library(tmap)
library(here)
library(tidyverse)
library(tidyr)
library(dplyr)
library(gt)
library(stars)
library(terra)
library(kableExtra)
library(spData)
library(spDataLarge)
library(geodata)
library(testthat)
library(leaflet)
library(raster)
```
</details>

## Function Code

To create a workflow to find the suitable areas based on two criteria: ocean depth and water temperature, a function was created with the following procedures:

- Verification of dataset coordinate system
- Process data: calculate mean temperature, convert temperature units, crop & resample raster to match all dataset extend and resolution
- Calculate suitable area: reclassify depth and temperature data based on defined criteria, combine rasters
- Create map

<details>
<summary>Click to view code</summary>
```{r}
# Define the suitable_habitat_func function
suitable_habitat_func <- function(tempt_combined,
                                  depth, 
                                  wc_regions,
                                  preferred_tempt_high,
                                  preferred_tempt_low,
                                  preferred_depth_high,
                                  preferred_depth_low) {

  #Step 1: Verify coordinate reference systems
  
  # List of datasets to check and reproject if necessary
  # Create a list of dataset names as strings
  dataset_names <- c("wc_regions", "tempt_combined", "depth")
  # Use mget to retrieve the datasets from the global environment
  datasets <- mget(dataset_names)

  # Loop over the datasets starting from the second dataset (index 2)
  for (i in 2:length(datasets)) {
    dataset_name <- datasets[[i]]
    # Check if the CRS of the current dataset matches the CRS of the first dataset
    if (crs(datasets[[i]]) == crs(wc_regions)) {
      print(paste("File",i, "CRS matches", names(datasets)[1]))
    } else {
      print(paste("File",i," CRS does not match, reprojecting"))
      # Reproject the dataset to match the CRS of the first dataset
      datasets[[i]] <- project(datasets[[i]], crs(wc_regions))
      #depth<-project(depth, crs(wc_regions))
      }
    }
  
  # Step 2: Process data
  
  # Calculate the mean temperature
  tempt_avg <- mean(tempt_combined, na.rm = TRUE)
  
  #Convert temperature from Kelvin to Celsius
  tempt_avg_c <- tempt_avg - 273.15
  
  # Crop the depth raster to match the extent of the tempt raster
  depth_cropped <- crop(depth, ext(tempt_combined))
  
  # Resample the depth raster to match the resolution of the tempt raster
  depth_resampled <- resample(depth_cropped, tempt_combined, method = "near")
  
  # Ensure rasters have the resolution, and CRS
  crs(depth_resampled) <- crs(depth)
  ext(depth_resampled) <- ext(tempt_combined)
  
  # Double check with conditional statements
  if (crs(tempt_combined) == crs(depth_resampled) && 
      all(ext(tempt_combined) == ext(depth_resampled)) && 
      all(res(tempt_combined) == res(depth_resampled))) {
    print("The rasters match in CRS, extent, and resolution.")
  } else {
    print("The rasters do not match in one or more aspects.")
  }
  
  # Step 3: Find suitable locations
  
  # Stack the rasters
  stacked_rasters <-c(tempt_avg_c, depth_resampled)
  
  # Reclassify temperature data
  rcl_tempt <- matrix(c(
    -Inf, preferred_tempt_low, 0,           
    preferred_tempt_low, preferred_tempt_high, 1,           
    preferred_tempt_high, Inf, 0),
    ncol = 3, byrow = TRUE)
  
  # Reclassify the mean temperature raster
  sst_reclassified <- classify(stacked_rasters$mean, rcl = rcl_tempt)
  
  # Reclassify depth: 
  rcl_depth <- matrix(c(
    -Inf, preferred_depth_low, 0,          
    preferred_depth_low, preferred_depth_high, 1,           
    preferred_depth_high, Inf, 0), 
    ncol = 3, byrow = TRUE)
  
  # Reclassify the depth raster
  depth_reclassified <- classify(stacked_rasters$depth, rcl = rcl_depth)
  
  # Stack the rasters
  suitable_locations <-c(sst_reclassified, depth_reclassified)
  
  # Combine the reclassified temperature and depth rasters
  suitable_locations <- lapp(suitable_locations, 
                                  fun = function(x, y) { return(x * y) })
  
  # Step 4: Create map
  
  # Create transparent color to hide not-suitable areas
  transparent <- rgb(1, 0, 0, alpha = 0)
  
  # Create an interactive map
  map <- leaflet() %>%
    addTiles() %>%  
    addRasterImage(suitable_locations, 
                   colors = c(transparent, "firebrick"),  
                   opacity = 0.6) %>%  
    addLegend(position = "topright", 
              pal = colorFactor(palette = c(transparent, "firebrick"), domain = c(1)),
              values = c(1),
              title = "Suitable Areas",
              labels = c("Suitable"),
              opacity = 1) 
  
  # Step 5: Determine the most suitable EEZ and compute total area
  
  # Rasterize the EEZ shapefile to match the resolution of the suitable locations raster
  eez_raster <- rasterize(wc_regions, suitable_locations, field = "rgn")
  
  # Mask the suitable locations raster by the EEZ regions (keep valid areas only)
  suitable_in_eez <- mask(suitable_locations, eez_raster)
  
  # Remove NAs
  suitable_in_eez[is.na(suitable_in_eez)] <- 0
  
  # Verify the results with a unit test**
  
  print("Executing test: does suitable_in_eez variable contains only 0 or 1?")
  # Test if suitable_in_eez only contains 0 or 1 (no NAs)
  test_that("suitable_in_eez variable contains only 0 or 1", {
    expect_true(all(suitable_in_eez[] %in% c(0, 1)))
  })
  
  # Calculate sum of suitable location for each region
  suitable_in_eez2 <- terra::zonal(suitable_in_eez,eez_raster , fun = "sum")
  suitable_in_eez2
  
  # Rename the columns to reflect the calculation (old label comes from the mean temperature calculation)
  colnames(suitable_in_eez2) <- c("mean", "suitable_sum")
  
  # Generate cell size summary
  cell_size <-cellSize(suitable_in_eez)
  
  # Reproject raster to UTM Zone 10N (EPSG:32610)
  # lon/lat WGS 84 is an angular measurement , we need metric unit
  projected_raster <- terra::project(suitable_in_eez, "EPSG:32610")

  # Extract area in meters (m²) from UTM projection
  cell_area_m2 <- cellSize(projected_raster)  
  cell_area_m2<- minmax(cell_area_m2)[1]

  # Print cell area for verification
  print(paste("Area of one cell (m2): ", cell_area_m2))
  
  # Calculate the total area for each region
  suitable_in_eez2<-suitable_in_eez2%>%
    mutate(total_area_m2=suitable_sum* cell_area_m2)
  
  # Step 6: Create summary table 
  
  # Using library gt to create a table to better present the results
  gt_table <- suitable_in_eez2 %>%
    gt() %>%
    tab_header(
      title = "Suitable Habitat Area by Region in California"
    ) %>%
    cols_label(
      mean = "Region",
      suitable_sum = "Number of Cells",
      total_area_m2="Total Area (m²)"
    ) %>%
    fmt_number(
      columns = c(total_area_m2),
      decimals = 0)

  # Assign the map and table to the global environment
  assign("map", map, envir = .GlobalEnv)
  assign("gt_table", gt_table, envir = .GlobalEnv)
}
``` 
</details>

## Load Data and Assign Input Variables

This section is to assign the filepath to read the data, define input variables:

- High Temperature Threshold: 22.3 C
- Low Temperature Threshold: 14.8C
- High Depth Threshold : 0
- Low Depth Threshold: -150 m


<details>
<summary>Click to view code</summary>
```{r}
# Load economic zones
wc_regions <- invisible(st_read(here::here("data", "wc_regions_clean.shp")))

# Load bathymetry
depth <- rast(here::here("data","depth.tif"))
depth<-project(depth, crs(wc_regions))

# Define the specie's ideal temperature and ocean depth
preferred_tempt_high<-22.3
preferred_tempt_low<-14.8
preferred_depth_high<-0
preferred_depth_low <- -150

# Define the temperature raster start and end year
tempt_year1 <- 2008
tempt_year2 <- 2012

# Loop over the years to load the temperature raster files
# Create an empty list to store the rasters
tempt_rasters <- list()
  
for (year in tempt_year1:tempt_year2) {
    
  # Create file name for each year
  file_name <- paste0("average_annual_sst_", year, ".tif")
    
  # Load raster
  tempt_rasters[[year]] <- rast(here::here("data", file_name))  
  }
  
# Combine all the rasters into one
tempt_combined <- rast(unlist(tempt_rasters, use.names = TRUE))

# Project the temperature combined dataset
tempt_combined <- project(tempt_combined, crs(wc_regions))
```
</details>

## Finding Suistainable Lobster Habitat Analysis

Section that runs the `suitable_habitat_func` function to obtain results:

<details>
<summary>Click to view code</summary>

```{r}
suitable_habitat_func (tempt_combined,
                       depth,
                       wc_regions,preferred_tempt_high,
                       preferred_tempt_low,
                       preferred_depth_high,
                       preferred_depth_low)
```

</details>

## California Map

The map below, depicts the areas in red where it meets both criteria: suitable depth and temperature.
```{r}
map
```

The following table summarizes the calculated total area by region where lobsters can thrive.

```{r}
gt_table
```

## Study Findings

The analysis presented above identifies the most suitable environmental conditions for a marine species in California, with a focus on prioritizing potential aquaculture locations. In this example, the species in question is the lobster, with the following optimal growth conditions:

Temperature: 14.8°C - 22.3°C
Depth: 0 - 150 meters

To conduct this analysis, Sea Surface Temperature and Bathymetry data were processed to assess the suitability of various regions. The function developed for this analysis computes the total suitable area for each region in California, taking into account both temperature and depth criteria.

Based on the results, the analysis suggests that Southern California offers the most favorable conditions for lobster growth, as both temperature and depth requirements are met in this region. This finding highlights Southern California as a prime location for lobster aquaculture, given its optimal environmental conditions.
