---
title: "California Spiny Lobster: Habitat Assessment for Aquaculture Opportunities"
author: "Karol Paya"
date: Dec. 7, 2024
execute: 
  eval: True
  warning: False
  message: False
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Study Background

This study employs a geospatial analysis to assess the suitability of habitats for spiny lobsters *(Panulirus interruptus)* in California, with a particular focus on identifying areas most conducive to aquaculture development. Spiny lobsters are an economically significant species in the region, and their population dynamics are influenced by factors such as temperature and ocean depth. However, traditional fishing practices have led to pressure on wild populations, prompting a growing interest in sustainable aquaculture as a potential solution. By remote sensing technologies, this analysis integrates environmental data, such as sea surface temperature and Bathymetry to search for potential areas that could support viable lobster farming operations.

Suitable Growing Conditions:

To identify the optimal habitat for California spiny lobsters, the following environmental criteria are considered, based on existing research:

-   Preferred Temperature Range: 14.8°C - 22.3°C

-   Preferred Depth Range: 0 - 150 meters

Source: [Sealifebase](https://www.sealifebase.ca/summary/Panulirus-interruptus.html), 2024

# Data Details

1. Sea Surface Temperature:

The Sea Surface Temperature (SST) dataset from NOAA (2008-2012) provides daily global satellite-derived sea surface temperature anomalies at a 5 km resolution. This dataset is crucial for identifying areas with suitable temperature ranges for spiny lobsters, as their growth and habitat preferences are linked to specific water temperature conditions (14.8°C - 22.3°C). It is used to reclassify temperature data and pinpoint suitable lobster farming areas.

2. Bathymetry:

The General Bathymetric Chart of the Oceans (GEBCO) providesbathymetric data that describes the ocean's depth. This dataset is essential for evaluating the ocean depth of potential lobster habitats, as spiny lobsters prefer specific depth ranges (0-150 meters).

3. Exclusive Economic Zones (EEZs):

The Exclusive Economic Zones (EEZs) dataset from MarineRegions.org defines the marine boundaries of countries.This dataset is used to mask the suitable lobster habitat areas by region, ensuring that only locations within the jurisdictional boundaries of relevant EEZs are considered in the final suitability analysis.


## References:

**Sea Surface Temperature**

National Oceanic and Atmospheric Administration (NOAA). 2008-2012. 5km Daily Global Satellite Sea Surface Temperature Anomaly v3.1 \[Data set\]. NOAA National Centers for Environmental Information. Retrieved Nov. 14, 2024 from https://www.ncei.noaa.gov

**Bathymetry**

General Bathymetric Chart of the Oceans (GEBCO). 2024. gridded bathymetry data \[Data set\]. Retrieved Nov. 14, 2024 from https://www.gebco.net/data_and_products/gridded_bathymetry_data/#area

**Exclusive Economic Zones**

MarineRegions.org. (n.d.). Exclusive Economic Zones (EEZs) off the west coast of the United States \[Data set\]. Retrieved Nov. 14, 2024 from https://www.marineregions.org/


# Geospatial Assessment: Finding suitable areas for lobster

## Load Libraries

<details>

<summary>Click to view code</summary>

```{r}
# Load libraries
library(tmap)
library(here)
library(tidyverse)
library(tidyr)
library(dplyr)
library(gt)
library(stars)
library(terra)
library(kableExtra)
library(spData)
library(spDataLarge)
library(geodata)
library(testthat)
library(leaflet)
library(raster)
```

</details>

## Creating a Function to Assess Suitable Areas

To identify the most suitable areas for spiny lobster aquaculture based on ocean depth and water temperature, a function was developed with the following steps:

1.  Verify Dataset Coordinate System

Ensuring all datasets (bathymetry and temperature) are using a consistent and compatible coordinate reference system to facilitate accurate spatial analysis.

2.  Process Data

    -   Calculate Mean Temperature: Compute the average temperature across the study area to account for temporal variability in ocean temperatures.

    -    Convert Temperature Units: Standardize temperature data by converting all values to a consistent unit (e.g., Celsius) for uniform analysis.

    -   Crop and Re-sample Rasters: Clip and re-sample the temperature and bathymetry rasters to match the same spatial extent and resolution, ensuring alignment for further analysis.

3.  Calculate Suitable Areas

    -   Reclassify Depth and Temperature Data: Based on predefined suitability criteria (e.g., preferred depth range: 0-150 m, preferred temperature range: 14.8°C - 22.3°C), reclassify both the bathymetric and temperature rasters into binary suitability maps (suitable/unsuitable).

    -   Combine Rasters: Overlay and combine the reclassified rasters to identify areas that meet both the depth and temperature requirements for spiny lobster habitat suitability.

4.  Create Final Map Generate a visual representation of the suitable areas, highlighting regions that fulfill the specified environmental criteria for potential aquaculture development.

<details>

<summary>Click to view code</summary>


```{r}
# Define the suitable_habitat_func function
suitable_habitat_func <- function(tempt_combined,
                                  depth, 
                                  wc_regions,
                                  preferred_tempt_high,
                                  preferred_tempt_low,
                                  preferred_depth_high,
                                  preferred_depth_low) {

  #Step 1: Verify coordinate reference systems
  
  # List of datasets to check and reproject if necessary
  # Create a list of dataset names as strings
  dataset_names <- c("wc_regions", "tempt_combined", "depth")
  # Use mget to retrieve the datasets from the global environment
  datasets <- mget(dataset_names)

  # Loop over the datasets starting from the second dataset (index 2)
  for (i in 2:length(datasets)) {
    dataset_name <- datasets[[i]]
    # Check if the CRS of the current dataset matches the CRS of the first dataset
    if (crs(datasets[[i]]) == crs(wc_regions)) {
      print(paste("File",i, "CRS matches", names(datasets)[1]))
    } else {
      print(paste("File",i," CRS does not match, reprojecting"))
      # Reproject the dataset to match the CRS of the first dataset
      datasets[[i]] <- project(datasets[[i]], crs(wc_regions))
      #depth<-project(depth, crs(wc_regions))
      }
    }
  
  # Step 2: Process data
  
  # Calculate the mean temperature
  tempt_avg <- mean(tempt_combined, na.rm = TRUE)
  
  #Convert temperature from Kelvin to Celsius
  tempt_avg_c <- tempt_avg - 273.15
  
  # Crop the depth raster to match the extent of the tempt raster
  depth_cropped <- crop(depth, ext(tempt_combined))
  
  # Resample the depth raster to match the resolution of the tempt raster
  depth_resampled <- resample(depth_cropped, tempt_combined, method = "near")
  
  # Ensure rasters have the resolution, and CRS
  crs(depth_resampled) <- crs(depth)
  ext(depth_resampled) <- ext(tempt_combined)
  
  # Double check with conditional statements
  if (crs(tempt_combined) == crs(depth_resampled) && 
      all(ext(tempt_combined) == ext(depth_resampled)) && 
      all(res(tempt_combined) == res(depth_resampled))) {
    print("The rasters match in CRS, extent, and resolution.")
  } else {
    print("The rasters do not match in one or more aspects.")
  }
  
  # Step 3: Find suitable locations
  
  # Stack the rasters
  stacked_rasters <-c(tempt_avg_c, depth_resampled)
  
  # Reclassify temperature data
  rcl_tempt <- matrix(c(
    -Inf, preferred_tempt_low, 0,           
    preferred_tempt_low, preferred_tempt_high, 1,           
    preferred_tempt_high, Inf, 0),
    ncol = 3, byrow = TRUE)
  
  # Reclassify the mean temperature raster
  sst_reclassified <- classify(stacked_rasters$mean, rcl = rcl_tempt)
  
  # Reclassify depth: 
  rcl_depth <- matrix(c(
    -Inf, preferred_depth_low, 0,          
    preferred_depth_low, preferred_depth_high, 1,           
    preferred_depth_high, Inf, 0), 
    ncol = 3, byrow = TRUE)
  
  # Reclassify the depth raster
  depth_reclassified <- classify(stacked_rasters$depth, rcl = rcl_depth)
  
  # Stack the rasters
  suitable_locations <-c(sst_reclassified, depth_reclassified)
  
  # Combine the reclassified temperature and depth rasters
  suitable_locations <- lapp(suitable_locations, 
                                  fun = function(x, y) { return(x * y) })
  
  # Step 4: Create map
  
  # Create transparent color to hide not-suitable areas
  transparent <- rgb(1, 0, 0, alpha = 0)
  
  # Create an interactive map
  map <- leaflet() %>%
    addTiles() %>%  
    addRasterImage(suitable_locations, 
                   colors = c(transparent, "firebrick"),  
                   opacity = 0.6) %>%  
    addLegend(position = "topright", 
              pal = colorFactor(palette = c(transparent, "firebrick"), domain = c(1)),
              values = c(1),
              title = "Suitable Areas",
              labels = c("Suitable"),
              opacity = 1) 
  
  # Step 5: Determine the most suitable EEZ and compute total area
  
  # Rasterize the EEZ shapefile to match the resolution of the suitable locations raster
  eez_raster <- rasterize(wc_regions, suitable_locations, field = "rgn")
  
  # Mask the suitable locations raster by the EEZ regions (keep valid areas only)
  suitable_in_eez <- mask(suitable_locations, eez_raster)
  
  # Remove NAs
  suitable_in_eez[is.na(suitable_in_eez)] <- 0
  
  # Verify the results with a unit test**
  
  print("Executing test: does suitable_in_eez variable contains only 0 or 1?")
  # Test if suitable_in_eez only contains 0 or 1 (no NAs)
  test_that("suitable_in_eez variable contains only 0 or 1", {
    expect_true(all(suitable_in_eez[] %in% c(0, 1)))
  })
  
  # Calculate sum of suitable location for each region
  suitable_in_eez2 <- terra::zonal(suitable_in_eez,eez_raster , fun = "sum")
  suitable_in_eez2
  
  # Rename the columns to reflect the calculation (old label comes from the mean temperature calculation)
  colnames(suitable_in_eez2) <- c("mean", "suitable_sum")
  
  # Generate cell size summary
  cell_size <-cellSize(suitable_in_eez)
  
  # Reproject raster to UTM Zone 10N (EPSG:32610)
  # lon/lat WGS 84 is an angular measurement , we need metric unit
  projected_raster <- terra::project(suitable_in_eez, "EPSG:32610")

  # Extract area in meters (m²) from UTM projection
  cell_area_m2 <- cellSize(projected_raster)  
  cell_area_m2<- minmax(cell_area_m2)[1]

  # Print cell area for verification
  print(paste("Area of one cell (m2): ", cell_area_m2))
  
  # Calculate the total area for each region
  suitable_in_eez2<-suitable_in_eez2%>%
    mutate(total_area_m2=suitable_sum* cell_area_m2)
  
  # Step 6: Create summary table 
  
  # Using library gt to create a table to better present the results
  gt_table <- suitable_in_eez2 %>%
    gt() %>%
    tab_header(
      title = "Suitable Habitat Area by Region in California"
    ) %>%
    cols_label(
      mean = "Region",
      suitable_sum = "Number of Cells",
      total_area_m2="Total Area (m²)"
    ) %>%
    fmt_number(
      columns = c(total_area_m2),
      decimals = 0)

  # Assign the map and table to the global environment
  assign("map", map, envir = .GlobalEnv)
  assign("gt_table", gt_table, envir = .GlobalEnv)
}
```

</details>

## Load Data & Define Criteria Variables

In order to perform the analysis and run the function for identifying suitable areas for lobster aquaculture, several key parameters need to be defined. 
First, the file paths must be specified. These include the path to the raster files containing water temperature and bathymetry data, which will be used to assess habitat suitability.

Additionally, the time period must be assigned to capture relevant environmental conditions over time. criteria variables for both temperature and depth also need to be defined, these are:

-   High Temperature Threshold: 22.3 C
-   Low Temperature Threshold: 14.8C
-   High Depth Threshold : 0
-   Low Depth Threshold: -150 m


<details>

<summary>Click to view code</summary>

```{r}
# Load economic zones
wc_regions <- invisible(st_read(here::here("data", "wc_regions_clean.shp")))

# Load bathymetry
depth <- rast(here::here("data","depth.tif"))
depth<-project(depth, crs(wc_regions))

# Define the specie's ideal temperature and ocean depth
preferred_tempt_high<-22.3
preferred_tempt_low<-14.8
preferred_depth_high<-0
preferred_depth_low <- -150

# Define the temperature raster start and end year
tempt_year1 <- 2008
tempt_year2 <- 2012

# Loop over the years to load the temperature raster files
# Create an empty list to store the rasters
tempt_rasters <- list()
  
for (year in tempt_year1:tempt_year2) {
    
  # Create file name for each year
  file_name <- paste0("average_annual_sst_", year, ".tif")
    
  # Load raster
  tempt_rasters[[year]] <- rast(here::here("data", file_name))  
  }
  
# Combine all the rasters into one
tempt_combined <- rast(unlist(tempt_rasters, use.names = TRUE))

# Project the temperature combined dataset
tempt_combined <- project(tempt_combined, crs(wc_regions))
```

</details>

## Finding Suistainable Lobster Habitat Analysis

In this section, we run the `suitable_habitat_func` function to identify and assess potential areas for sustainable spiny lobster aquaculture.  

<details>

<summary>Click to view code</summary>

```{r}
# Run function
suitable_habitat_func (tempt_combined,
                       depth,
                       wc_regions,preferred_tempt_high,
                       preferred_tempt_low,
                       preferred_depth_high,
                       preferred_depth_low)
```

</details>

## Suitable Area for Lobster Aquaculture Map

The map below illustrates the areas (highlighted in red) that meet both the depth and temperature criteria for spiny lobster aquaculture. The red areas on the map indicate zones where both the ocean depth (0-150 meters) and water temperature (14.8°C - 22.3°C) fall within the preferred ranges for sustainable aquaculture.

```{r}
# Plot map
map
```

The following table summarizes the calculated total area by region where lobsters can thrive.

```{r}
# Display Summary Table
gt_table
```

## Study Findings

The analysis presented above identifies the most suitable environmental conditions for a marine species in California, with a focus on prioritizing potential aquaculture locations. In this example, the species in question is the lobster, with the following optimal growth conditions:

Temperature: 14.8°C - 22.3°C 

Depth: 0 - 150 meters

To conduct this analysis, Sea Surface Temperature and Bathymetry data were processed to assess the suitability of various regions. The function developed for this analysis computes the total suitable area for each region in California, taking into account both temperature and depth criteria.

Based on the results, the analysis suggests that Southern California offers the most favorable conditions for lobster growth, as both temperature and depth requirements are met in this region. This finding highlights Southern California as a prime location for lobster aquaculture, given its optimal environmental conditions.
